Synthesizing Quotient Lenses

### 2018-02-16 papers/QuotientLenses/main.tex
intro edits

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{boomerang} is the only implementation
currently offering quotient lenses. Even though quotient lenses permit simpler
transformations between the formats, they merely push complexity elsewhere. To
write a quotient lens, programmers must write with standard lens
combinators and \emph{canonizers}. Canonizers encode two functions, one that
maps a format to a canonical version of that format, and another that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to enforce that a quotient lens maps
between two different formats, they must write regular expressions specifying
such formats. In this way, writing quotient lenses involves writing 3 terms: (1)
the data formats on each side of the lens, (2) canonizers to normalize the data
and choose representatives, and (3) a lens between the data formats when in
canonical form, even though, typically, these three share a great deal of
structure. Moreover, writing the lens between the equivalence classes can be
challenging due to the rather fiddly typechecking, mainly due to
``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we make programming quotient lenses easier in two ways. First, we
introduce a novel language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression as well as
a set of canonical representatives on the regular expression's language. With
QREs, programmers do not need to define the data and canonization functions
separately -- by adding simple annotations to the data, the canonization
functions can be inferred.

Second, we use QREs to define a class of quotient lenses called {\em QRE
lenses} which map betweeen QREs, and show how to synthesize QRE lenses. In
order to derive a lens, a programmer need not manually derive the lens
herself, she only needs to provide a pair of QREs describing the formats the
lens should map between, and a suite of examples that demonstrates how the
lens should act on specific examples.  The resulting lens is guaranteed to
map between the languages of the QREs and satisfy the provided examples.

In summary, our main contributions are:
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs)
  which provide a compact, convenient notation that expresses an equivalence
  relation on a regular languages as well as as a canonical set representatives
  for this equavalence relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}, a subset of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution in this
  paper is a proof that each QRE lens $q$ defines the same transformation as a
  QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
  \item Using this normal form, we reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate its applicability by using it to
  synthesize QRE lenses between several real-world data formats from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

>>>>

%However, while plain lenses are widely used in both standalone systems and
%popular libraries, Boomerang~\cite{boomerang} is the only implementation
%currently offering quotient lenses. Even though quotient lenses permit simpler
%transformations between the formats, they merely push complexity elsewhere.
Ordinary bijective lenses are fairly tricky to program, especially for
beginners, as they require programmers to think in two directions at once
and learn a new type system, which comes with some tricky ambiguity constraints.
Unfortunately, quotient lenses do not improve the situation:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we improve the process of programming quotient lenses
in two distinct, but connected ways. First, we
introduce a new language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression as well as
a set of canonical representatives for strings that match the regular
expression. With
QREs, programmers do not need to define the data and canonization functions
separately -- by adding simple annotations to ordinary regular expressions,
canonization functions can be inferred.

Second, we use QREs to define a class of quotient lenses called {\em QRE
lenses}, and, we show how to synthesize such QRE
lenses automatically.  
More specifically, in
order to generate a lens, a programmer provides a pair of QREs
describing the source and target formats, and, optionally,
a suite of examples that demonstrates how the
lens should act on specific examples.  Given this information,
we are able to exploit and extend recent work on algorithms for synthesis of
bijective lenses~\cite{optician} to generate quotient lenses.
Importantly, the lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (a fact that cuts down the amount of non-determinism in
a specification and thus helps speed up synthesis), we prove that
this rigidity does not compromise expressiveness.

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs)
  which provide a compact, convenient notation that expresses an equivalence
  relation on a regular languages as well as a canonical set representatives
  for this equavalence relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that each QRE lens $q$ defines the same transformation as a
  QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
  \item Using this normal form, we reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between several real-world data formats from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.
